Impacts of Month, Grocery Vendor, Presale Prices, and Historical Pricing on Discounted Egg Dozen Prices

### 2024-12-04 paper/paper.qmd
added discussion + edited readme

---
title: "Impacts of Month, Grocery Vendor, and Presale Prices on Discounted Egg Dozen Prices"
subtitle: "Cheapest Discounted Prices Are Offered by T&T in April, by Multiple Linear Regression"
author: Cher Ning-Li
thanks: "Code and data are available at: (https://github.com/cher-ning/cndgrocers)[https://github.com/cher-ning/cndgrocers]."
date: today
date-format: long
abstract: "This study aims to understand the impacts of month, vendor, presale price, and average historical item pricing on the current discounted price of egg dozens using data of sales held between March and November 2024 by the 8 major grocery vendors of Canada. Using a multiple linear regression model, these factors were found to have significant interactions with the discounted price of eggs. These findings are key to understanding pricing patterns of different vendors across different months, helping consumers and retailers make better informed decisions."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Evidently, the price of eggs has crucial impact on the lives of a huge number of people, whether it be all the consumers who are purchasing eggs regularly, or the farmers who make their living by selling this product. Between these two parties is the vendors, who determine the price at which these crucial food items are sold to customers for. Therefore, it would be beneficial to understand the underlying factors that determine egg pricing at various vendors.

@@

## Variables of Interest

@@

For the continuous variables `old_price`, we observe a negative coefficient of $-0.80$, which may be surprising at first because one would expect discounted price to increase as presale price increases. However, after taking the intercept base price into consideration, the interpretation for this coefficient makes much more sense. This result indicates that given a set base price, a greater presale price is correlated to a lower discounted price.

@@

The $R^2$ and Adjusted $R^2$ values of 0.82 indicate that $82%$ of the variability in the discounted price within this dataset can be explained by the current model.

```{r}
#| echo: false
#| eval: true
#| label: fig-vendorresults
#| fig-cap: "Vendor Category Coefficients"
#| warning: false

@@

price_permonth |> 
  ggplot(mapping = aes(x = month, y = avg_price)) +
  geom_col() + 
  theme_classic() +
  labs(x = "Month",
       y = "Average Sale Price of a Dozen Eggs")

@@

First, we compare the vendor categories in more detail in @fig-vendorresults. `vendorTandT` has the lowest coefficient value at $-1.87$, meaning that the average discounted price at T&T is expected to be the lowest, compared to other vendors. `vendorSaveOnFoods` has the highest coefficient at $0.02$, is it also the only positive coefficient for the vendors, meaning that average discounted prices at Save On Foods are expected to be the highest, and the only vendor which has higher average prices than the baseline category of Galleria.

```{r}
#| echo: false
#| eval: true
#| label: fig-monthresults
#| fig-cap: "Month Category Coefficients"
#| warning: false

@@

## Factors Influencing Egg Dozen Prices {#sec-influencing-factors}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

>>>>

---
title: "Impacts of Month, Grocery Vendor, Presale Prices, and Historical Pricing on Discounted Egg Dozen Prices"
subtitle: "Cheapest Discounted Prices Are Offered by T&T or in April, by Multiple Linear Regression"
author: Cher Ning-Li
thanks: "Code and data are available at: [https://github.com/cher-ning/cndgrocers](https://github.com/cher-ning/cndgrocers)."
date: today
date-format: long
abstract: "This study aims to understand the impacts of month, vendor, presale price, and average historical item pricing on the current discounted price of egg dozens using data of sales held between March and November 2024 by the 8 major grocery vendors of Canada. Using a multiple linear regression model, these factors were found to have significant interactions with the discounted price of eggs, with T&T and April observing the lowest prices within each of their own categories. These findings are key to understanding pricing patterns of different vendors across different months, helping consumers and retailers make better informed decisions."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Evidently, the price of eggs has crucial impact on the lives of a huge number of people, whether it be all the consumers who are purchasing eggs regularly, or the farmers who make their living by selling this product. Between these two parties is the vendors, who determine the price at which these staple food items are sold to customers for. Therefore, it would be beneficial to understand the underlying factors that determine egg pricing at various vendors.

@@

## Variables of Interest {#sec-variables}

@@

For the continuous variable `old_price`, we observe a negative coefficient of $-0.80$, indicating that when all other variables are held constant, a greater presale price is correlated with a decrease in current discounted price. Intuitions behind why this value may be negative will be discussed in @sec-discussion.

@@

The $R^2$ and Adjusted $R^2$ values of 0.82 indicate that 82% of the variability in the discounted price within this dataset can be explained by the current model.

```{r}
#| echo: false
#| eval: true
#| label: fig-vendors-result
#| fig-cap: "Coefficients Representing Each of the 8 Vendors' Impact on Discounted Price"
#| warning: false

@@

@@

First, we compare the vendor categories in more detail in @fig-vendors-result. Variable `vendorTandT` has the lowest coefficient value at $-1.87$, meaning that the average discounted price at T&T is expected to be the lowest, compared to other vendors. `vendorSaveOnFoods` has the highest coefficient at $0.02$, is it also the only positive coefficient for the vendors, meaning that average discounted prices at Save On Foods are expected to be the highest, and the only vendor which has higher average prices than the baseline category of Galleria.

```{r}
#| echo: false
#| eval: true
#| label: fig-monthresults
#| fig-cap: "Coefficients Representing Each Month's Impact on Discounted Price From April to November"
#| warning: false

@@

## Main Findings and Implications {#sec-influencing-factors}

The multiple linear regression model built in this paper investigates the influences of month, vendor, presale pricing, and average discounted price of the previous month on current egg item discounts. This model is helpful in isolating out the effects of each variable on price. In the vendor category, the effects of each vendor translates quite directly into the average listed discount price, with T&T and No Frills having the lowest coefficients as well as average prices (@fig-vendorprices; @fig-vendors-result). With these results, it can concluded that when considered alone, the vendors where customers can go to find the lowest to highest discounted egg prices would be T&T, No Frills, Metro, Loblaws, Voila, Walmart, Galleria, and Save On Foods, in that order.

More interestingly, as shown in @fig-monthprices in @sec-variables, July appeared to be the month where lowest discounted prices are offered. However, @fig-monthresults show that after isolating out the effects of other variables, it is actually the month of April that contributes most to discounting prices compared to other months. This discrepancy points to other confounding factors at play that caused the July average prices to be lowest. Further investigation with consideration for possible alternate causes would be needed to understand this pattern better. In contrast to the patterns observed in the simple data overview shown previously (@fig-monthprices), the model results now indicate that when considered individually, the months associated with the lowest to highest discounted egg prices would be April, September, May, November, June, July, October, August, in that order.

The variable `old_price`, representing an item's presale price, was found to have a negative coefficient at $-0.80$. This may be surprising at first because one would expect discounted price to increase as presale price increases. However, after taking the intercept base price into consideration, the interpretation for this coefficient makes much more sense. This result indicates that given every other variable is held constant, a greater presale price is correlated to a lower discounted price. This understanding makes much more intuitive sense, as sale events are typically calculated by percentage of the original price, meaning that a greater original price would lead to a greater decrease in price, thereby causing a greater decrease in discounted price in the end.

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-resid-model
#| fig-cap: "Residual vs Fitted Values Plot for Model"

plot(model, which = 1)
```

@fig-resid-model shows the residuals vs fitted values for this paper's model. This plot can be used to check the linearity and independent error assumptions of Multiple Linear Regression modeling. Our result does not fully resemble a null plot, as expected under these assumptions, indicating that some non-linear transformations on the predictor variables, response variable, or some combination of both can be used to improve the fit of this model.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-qqplot
#| fig-cap: "QQ-Normal Plot for Model"

plot(model, which = 2)
```

@fig-qqplot shows the QQ plot for the current model, which checks for the normal error assumption of Linear Regression modelling. As observed, the points deviate heavily from the straight line on the lower end, indicating that errors are not normally distributed.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-standardizedfitted
#| fig-cap: "Standardized Residuals vs Fitted Values Plot for Model"

plot(model, which = 3)
```

@fig-standardizedfitted shows the standardized residuals vs fitted values graph for this model, which can be used to check the constant variance assumption of Linear Regression modelling. Our graph shows clustering, rather than an even spread around 0 that is expected, indicating a violation of this assumption. This means that a variance-stabilizing transformation on the response variable would be required to improve model accuracy further.
